---
title: "Working Markdown"
author: "Maryam Nouri-Aiin"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: journal
    highlight: monochrome
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For markdown, use asterisks to *indicate italic text*.

Double line return for a new chunk of text like this.

Name  
Department of Plant and Soil Science  
University of Vermont  


# Hash marks for headers


## More hashes

### Smaller

#### and smaller

##### headers!!


**bold face this way**  


>offset quote which embeds ans looks really nice when you extended material that you would like to see.  


Back to plain text  



* fist list item  
* second list item  
  - spacing for indents
  
  
    ^superscript^ but also a ~subscript~ lik this.  
    
[set up a link with text](website addrss)

also footnotes^[at bottom of page]


First Header  | Second Header
------------- | -------------
Content Cell  | Content Cell
Content Cell  | Content Cell

```
Fencing sets bounderies and everything inside the fencing *is shown as literal text* whitout rendering.
```

$$a +b =c$$

We can also use Latex in line with a single dollar sign $a +b=c$ like this.


## subscripts in LaTex

$$H_0 = Z_{a + b}$$

## subscripts in LaTex

$$S = cA^z$$

$$S=cA^z_1 + z_{2+x}$$
$$\alpha = \frac{\beta}{\delta + \gamma_x}$$

$$z= \sum_{i=1}^X{K}$$


## Escaping the special character in LaTex

$$\backslash \alpha \le b \backslash$$

## Rendering in plain text

$$P(Occurance of Species A)=Z$$
$$P(\mbox{Occurance of Species A})=Z$$

Easy to embded LATex in line $\pi + 3$ link so
We can also use makdown fencing in line with a single backtick `pi+3` like this. we can even excute r code inside of markdown text `r pi + 3`.

```{r eval=TRUE, echo=TRUE}
# Use comments extensively in ALL of your coding!
Pred <- seq(1,10)     # make a vector of integers from 1 to 10
Res <- runif(10)      # generate 10 draws from a random uniform (0,1) distribution

# print the random numbers
print(Res)

# plot the graph
plot(x=Pred,y=Res,type="b")
```
